How topos theory can help algebra and geometry

### 2018-07-10 slides-leipzig2018.tex
Finished the notes

\usepackage{mathtools}
\usepackage{booktabs}
\usepackage{stmaryrd}
\usepackage{array}
\usepackage{ragged2e}
\usepackage{multicol}
\usepackage{tabto}
\usepackage{xstring}
\usepackage{ifthen}
\usepackage{soul}\setul{0.3ex}{}
\usepackage[all]{xy}
\xyoption{rotate}
\usepackage{tikz}
\usetikzlibrary{mindmap,calc,shapes,shapes.callouts,shapes.arrows,patterns,fit,backgrounds,decorations.pathmorphing}
\hypersetup{colorlinks=false}

@@

  \scriptsize
  Ingo Blechschmidt \\
  July 10th, 2018
  \par
\end{frame}}

@@

  Toposes generalize topological spaces in two ways: Firstly, the ``open sets''
  of toposes don't actually have to be sets of points; they can be more general
  kinds of objects such as coverings. Toposes can even have no points at all
  and still be nontrivial (this is for instance the case for the \emph{topos of
  random sequences}). Secondly, while classically a given open
  subset is either contained in a further open subset or not, the opens of
  toposes can be contained in further opens in many different ways.
  % XXX: Link to Alex Simpson

  Recently, toposes are being used to help the mysterious field with
  one element come into being.}
  % XXX: Link to John Baez

@@

  Crucially, two theories~$T$ and~$T'$ can have equivalent classifying
  toposes even when they are not syntactially related in any way. This
  observation is the starting point of Olivia Caramello's \emph{bridge
  technique}, a grand research program with applications in many different
  fields.
  % XXX: Link Olivia

@@

  \jnote{2}{Besides~$\Set$, there is a proper class worth of further toposes.
  We'll get to know some of these better during the course of the
  talk.

@@

  Most of the toposes in active use are either toposes of sheaves (over a
  topological space or a XXXnLab \emph{site}), realizability toposes (such
  as~$\Eff$ or variants constructed using different models of computation), or
  arise from those using topos-theoretic constructions; but this is not a
  complete classification.}
\end{frame}

@@

  \jnote{2}{When exploring a new topos for the first time, the only way to
  find out which statements hold in it is to translate them using the
  Kripke--Joyal semantics and check whether the translation holds in the usual
  mathematical sense. As soon as we have established a certain Grundstock XXX,
  we can switch to a more efficient procedure: We just use mathematical
  reasoning to deduce from known statements new ones.}

@@

  \begin{enumerate}\justifying
    \item The usual proof that~$\sqrt{2}$ is not rational is perfectly fine
    from a constructive point of view. It shows that the assumption
    that~$\sqrt{2}$ is rational entails a contradiction. This is just the
    definition of what it means to be not rational. (There's a difference
    XXX Link Andrej
    between a true proof by contradiction and a proof of a negated statement.
    Only the former can't generally be carried out in constructive
    mathematics.)
    \item Constructively it's still true that there is no such thing as a
    statement being neither true nor false: That is, we still
    have~$\neg\neg(\varphi \vee \neg\varphi)$.
  \end{enumerate}}

@@

    Both the failure and the two fixes can be interpreted computationally:
    Given just the promise of an inhabited subset, we can't algorithmically
    determine its minimum. But we can do so when given a \emph{membership
    oracle}, or if it's okay to return the result in the
    \emph{continuation monad}. XXX Link Ingo

    \item We can't constructively prove that any finitely generated vector
    space admits a basis. We can, however, constructively verify that any
    finitely generated vector does \emph{not not} admit a basis. (By exploiting
    that given a generating family~$(x_1,\ldots,x_n)$ it's \emph{not not} the
    case that either one of the generators can be expressed as a linear
    combination of the others, or not.) We'll see below that this particular
    example implies that any sheaf of finite type over a reduced scheme is
    locally free on a dense open.
  \end{enumerate}}

@@

    \item Similarly, we can't constructively verify the Gelfand--Naimark
    correspondence between commutative~$C^*$-algebras with unit and compact Hausdorff
    spaces. But we can do so if we pass from compact Hausdorff spaces to
    compact Hausdorff locales.
    XXX Naimark
  \end{enumerate}

  A recommendation for more on constructive mathematics is the informative and
  entertaining talk \emph{Five Stages of Accepting Constructive Mathematics}
  by Andrej Bauer (video, notes). XXX Link}

@@

@@

  \jnote{1}{There is a variant of the effective topos which is not built using
  Turing machines, but using \emph{infinite-time Turing machines}, a popular
  model for hypercomputation. In that variant, the statement ``any
  function~$\NN \to \NN$ is the zero function or not'' is true; the statement
  ``any function~$\NN \to \NN$ is computable by a Turing machine'' is false;
  and the statement~``any function~$\NN \to \NN$ is computable by an
  infinite-time Turing machine'' is true again. Details can be found in
  \fixedhref{...}{this set of slides}. XXX

  \begin{center}\includegraphics[width=0.6\textwidth]{exploring-hypercomputation-slides}\end{center}}

  \jnote{2}{We can also build a variant of the effective topos which uses
  \emph{machines of the real world} instead of idealized Turing machines. In
  doing so, we leave the realm of rigorous mathematics, but obtain interesting
  connections with philosophy and physics. For instance, in that variant the
  statement~``any function~$\RR \to \RR$ is continuous'' is true if machines in
  the real world can only perform finitely many computational steps in finite
  time and if it's possible to build hidden communication channels. Details can
  be found in the book chapter \emph{\fixedhref{...}{Realizability for the Real
  World}} by Andrej Bauer.
  \begin{center}\includegraphics[width=0.3\textwidth]{zenil-computable-universe}\end{center}}
\end{frame}

@@

    \begin{column}{0.35\textwidth}
      \appl{bohr-topos}{further subjects}{
        synth.~computability th. \\
        synth.~measure theory \\
        Bohr topos for QM
      }
    \end{column}
  \end{columns}
\end{frame}

@@

    \item The approximative version of the intermediate value theorem (stating
    that for any~$\varepsilon > 0$, there is a number~$x$ such that~$|f(x)| <
    \varepsilon$) has a constructive proof and therefore holds in~$\Sh(X)$. The
    external interpretation is that in continuous families of continuous
    functions, approximate zeros can locally be picked continuously.
    % XXX Link Matt F.

@@

    \item Write~``$M^\Delta$'' for the set of all maps~$\Delta \to M$. A
    \emph{vector field} on a manifold (set)~$M$ is just a map~$X : M \to
    M^\Delta$ such that~$X(p)(0) = p$ for all~$p \in M$. A tangent field~$X$
    induces an infinitesimal path~$\gamma : \Delta \to M^M$ in the space (set)
    of transformations of~$M$, by setting~$\gamma(\varepsilon) = (p \mapsto
    X(p)(\varepsilon))$.

@@

  Synthetic differential geometry is very well-developed and at your disposal.
  References include:
  \begin{itemize}
    \item \fixedhref{...}{Notes for highschool students} (in German)
    \item \fixedhref{...}{...}, a blog post by Andrej Bauer
    \item \fixedhref{...}{...}, a short introduction by ... (Primer)
    \item \fixedhref{...}{...}, the definitive book by Anders Kock
  \end{itemize}

@@

\section[Commutative algebra]{Applications to commutative algebra}

@@

    \item The ring~$A$ is a Prüfer domain if and only if~$A^\sim$ is a Bézout
    domain. Therefore any constructive theorem about Bézout domains entails a
    corresponding theorem about Prüfer domains. Bézout domains are quite rare,
    while Prüfer domains abound (for instance the ring of integers of any
    number field is a Prüfer domain).
  \end{itemize}}

@@

  Details are \fixedhref{...}{here} (XXX link Munich slides).

  As mentioned before, any proof involving the internal language can be
  \emph{unwound} to yield a direct proof not referencing toposes. Depending on
  the logical complexity of the statements appearing in the internal proof,
  this process can substantially lengthen the proof. In the case of the new
  proof of Grothendieck's generic freeness lemma, we were lucky to
  \fixedhref{obtain a one-page proof}{...} using this process. (This crucially
  rested on the fact that we were able to eliminate the use of the Noetherian
  property from the internal proof; else the unwound proof would have been much
  longer.)}
\end{frame}

\backupstart

@@

@@

  \hil{Synthetic approach:} model schemes \hil{directly as sets} in
  the internal universe of the \mbox{\hil{big Zariski topos}} of a base scheme.
  \small
  \[ \{ (x,y,z) \? (\affl)^3 \,|\, x^n+y^n-z^n=0 \} \]
\end{frame}

\backupend

>>>>

\usepackage{mathtools}
\usepackage{booktabs}
\usepackage{stmaryrd}
\usepackage{array}
\usepackage{ragged2e}
\usepackage{multicol}
\usepackage{tabto}
\usepackage{xstring}
\usepackage{ifthen}
\usepackage{soul}\setul{0.3ex}{}
\usepackage[all]{xy}
\xyoption{rotate}
\usepackage{tikz}
\usetikzlibrary{mindmap,calc,shapes,shapes.callouts,shapes.arrows,patterns,fit,backgrounds,decorations.pathmorphing}
\hypersetup{colorlinks=true}

@@

  \scriptsize
  Ingo Blechschmidt \\
  July 10th, 2018
  \par

  \jnote{1}{This talk has two parts. In the first part, we'll learn that
  besides the standard mathematical universe, in which ordinary mathematics
  takes place, there is a host of alternate mathematical universes. In these
  alternate universes, the usual objects of mathematics enjoy slightly
  different properties. For instance, we'll encounter universes in which the
  intermediate value theorem fails or in which any map~$\RR \to \RR$ is
  continuous.

  In the second part, we'll see that these alternate universes, while seeming
  strange on first contact, yield concrete applications in algebra and
  geometry.}
\end{frame}}

@@

  Toposes generalize topological spaces in two ways: Firstly, the ``open sets''
  of toposes don't actually have to be sets of points; they can be more general
  kinds of objects such as coverings. Toposes can even have no points at all
  and still be nontrivial (this is for instance the case for the
  \fixedhref{https://homepages.inf.ed.ac.uk/als/Talks/ccc09.pdf}{\emph{topos of
  random sequences}}). Secondly, while classically a given open
  subset is either contained in a further open subset or not, the opens of
  toposes can be contained in further opens in many different ways.

  Recently, toposes are being used to help the mythical
  \fixedhref{http://math.ucr.edu/home/baez/week259.html}{field with
  one element} come into being.}

@@

  Crucially, two theories~$T$ and~$T'$ can have equivalent classifying
  toposes even when they are not syntactially related in any way. This
  observation is the starting point of Olivia Caramello's
  \fixedhref{https://www.oliviacaramello.com/}{\emph{bridge
  technique}}, a grand research program with applications in many different
  fields.

@@

  \jnote{2}{Besides~$\Set$, there is a proper class' worth of further toposes.
  We'll get to know some of these better during the course of the
  talk.

@@

  Most of the toposes in active use are either toposes of sheaves (over a
  topological space or a
  \fixedhref{https://ncatlab.org/nlab/show/site}{\emph{site}}), realizability
  toposes (such as~$\Eff$ or variants constructed using different models of
  computation), or arise from those using topos-theoretic constructions; but
  this is not a complete classification. (The topos~$\Set$ is equivalent to the
  topos of sheaves over the one-point space.)}
\end{frame}

@@

  \jnote{2}{When exploring a new topos for the first time, the only way to
  find out which statements hold in it is to translate them using the
  Kripke--Joyal semantics and check whether the translation holds in the usual
  mathematical sense. As soon as we have established a certain stock of
  statements in this way, we can switch to a more efficient procedure: We can
  just use mathematical reasoning to deduce new statements from known ones.}

@@

  \begin{enumerate}\justifying
    \item The usual proof that~$\sqrt{2}$ is not rational is perfectly fine
    from a constructive point of view. It shows that the assumption
    that~$\sqrt{2}$ is rational entails a contradiction. This is just the
    definition of what it means to be not rational.
    (\fixedhref{http://math.andrej.com/2010/03/29/proof-of-negation-and-proof-by-contradiction/}{There's a difference}
    between a proof by contradiction and a proof of a negated statement.
    Only the former can't generally be carried out in constructive
    mathematics.)
    \item Constructively it's still true that there is no such thing as a
    statement which is neither true nor false: That is, we still
    have~$\neg\neg(\varphi \vee \neg\varphi)$.
  \end{enumerate}}

@@

    Both the failure and the two fixes can be interpreted computationally:
    Given just the promise of an inhabited subset, we can't algorithmically
    determine its minimum. But we can do so when given a \emph{membership
    oracle}, or if it's
    \fixedhref{https://rawgit.com/iblech/talk-constructive-mathematics/master/hal2015-notes.pdf}{okay to return the result in the
    \emph{continuation monad}}.

    \item We can't constructively prove that any finitely generated vector
    space admits a basis. We can, however, constructively verify that any
    finitely generated vector does \emph{not not} admit a basis, (By exploiting
    that given a generating family~$(x_1,\ldots,x_n)$, it's \emph{not not} the
    case that either one of the generators is a linear combination of the
    others, or not.) We'll see below that this particular example entails that
    any sheaf of finite type over a reduced scheme is locally free on a dense
    open.
  \end{enumerate}}

@@

    \item Similarly, we can't constructively verify the Gelfand--Neumark
    correspondence between commutative~$C^*$-algebras with unit and compact Hausdorff
    spaces. But we can do so if we pass from compact Hausdorff spaces to
    compact Hausdorff locales.
  \end{enumerate}

  A recommendation for more on constructive mathematics is the informative and
  entertaining talk \emph{Five Stages of Accepting Constructive Mathematics}
  by Andrej Bauer
  (\fixedhref{https://www.youtube.com/watch?v=21qPOReu4FI}{video},
  \fixedhref{http://www.ams.org/journals/bull/2017-54-03/S0273-0979-2016-01556-4/}{notes}).}

@@

      \item \intex{
        $\Eff \models \text{``Any function~$\RR \to \RR$ is continuous.''}$
      }{\textcolor{green!90}{\cmark}}{}

@@

  \jnote{1}{There is a variant of the effective topos which is not built using
  Turing machines, but using \emph{infinite-time Turing machines}, a popular
  model for hypercomputation. In that variant, the statement ``any
  function~$\NN \to \NN$ is the zero function or not'' is true; the statement
  ``any function~$\NN \to \NN$ is computable by a Turing machine'' is false;
  and the statement~``any function~$\NN \to \NN$ is computable by an
  infinite-time Turing machine'' is true again. Details can be found in
  \fixedhref{https://rawgit.com/iblech/mathezirkel-kurs/master/superturingmaschinen/slides-warwick2017.pdf}{this set of slides}.

  \begin{center}\fixedhref{https://rawgit.com/iblech/mathezirkel-kurs/master/superturingmaschinen/slides-warwick2017.pdf}{\includegraphics[width=0.6\textwidth]{exploring-hypercomputation-slides}}\end{center}}

  \jnote{2}{We can also build a variant of the effective topos which uses
  \emph{machines of the real world} instead of idealized Turing machines. In
  doing so, we leave the realm of rigorous mathematics, but obtain interesting
  connections with philosophy and physics. For instance, in that variant the
  statement~``any function~$\RR \to \RR$ is continuous'' is true if machines in
  the real world can only perform finitely many computational steps in finite
  time and if it's possible to build hidden communication channels. Details can
  be found in the book chapter
  \emph{\fixedhref{http://math.andrej.com/2014/03/04/intuitionistic-mathematics-and-realizability-in-the-physical-world/}{Intuitionistic Mathematics and Realizability in the Physical World}} by Andrej Bauer.
  \begin{center}\fixedhref{http://math.andrej.com/2014/03/04/intuitionistic-mathematics-and-realizability-in-the-physical-world/}{\includegraphics[width=0.3\textwidth]{zenil-computable-universe}}\end{center}}
\end{frame}

@@

    \begin{column}{0.35\textwidth}
      \appl{bohr-topos}{further subjects}{
        synth.~computability th. \\
        synth.~measure theory \\
        Bohr topos for QM
      }
    \end{column}
  \end{columns}

  \jnote{1}{The applications of the internal language of toposes in several
  branches of mathematics are driven by the following fundamental empirical fact:
  \begin{center}The external translations of \\
  some easy-to-prove internal statements of suitable toposes \\
  are \\
  nontrivial statements about the subject matter at hand.\end{center}
  For instance, Grothendieck's generic freeness lemma, an important theorem in
  commutative algebra and algebraic geometry, is the external translation of
  the basic observation ``finitely generated modules over finitely generated
  algebras over fields are \emph{not not} free over the field and \emph{not
  not} finitely presented over the algebra'' of constructive linear algebra.

  Is there a theorem which can only be proven by topos-theoretic methods?
  \emph{No,} for any proof employing the internal language can be unwound to
  yield an external proof not referencing topos theory. Just as the translation
  from internal statements to external ones is entirely mechanical, so is the
  translation from internal proofs to external ones. However, this translation
  will always make the proof longer, and, depending on the
  syntactical complexity of the involved statements, the resulting external
  proof might be quite complex.

  The next slide shows two examples.}

  \jnote{2}{\begin{itemize}\justifying
    \item The statement ``if the two outer sheaves in a short exact sequence of
    sheaves of modules are of finite type, then so is the middle one'' is the
    external translation of the basic fact ``if the two outer modules
    in a short exact sequence of modules are finitely generated, then so is the
    middle one'' of constructive linear algebra.

    The syntactical complexity of the internal statement is quite low (just one
    implication sign, on toplevel, no long chains of quantifiers of mixed
    types). Therefore the resulting external proof obtained by unwinding will
    still be quite short and straightforward. In fact, this external proof will
    coincide with what anyone well-versed in scheme theory will produce,
    judiciously juggling with open subsets.

    Using the internal language of toposes here is therefore mostly for mental
    hygiene, making rigorous the intuitive idea that the two statements are
    related to each other.

    \item A version of Hilbert's basis theorem, ``polynomial rings over
    anonymously Noetherian rings are again anonymously Noetherian'', can be put
    to use in the proof of Grothendieck's generic freeness lemma alluded to
    above. This statement has nontrivial syntactical complexity (the
    Noetherian condition involves a double negation, therefore nested
    implications). For this reason, already the translation of the
    \emph{statement} is quite convoluted (a part of it is
    \fixedhref{https://rawgit.com/iblech/internal-methods/master/slides-como2018.pdf}{displayed
    in this set of slides} for your viewing pleasure). The translation of its
    \emph{proof} contains a deformed copy of the standard proof of Hilbert's
    basis theorem (in a way which doesn't allow to simply \emph{apply}
    Hilbert's basis theorem, one has to actually redo its proof).

    In situations like this, the internal language is of real value.
  \end{itemize}}

  \jnote{3}{We're used to the fact that the usual laws of mathematical
  reasoning apply to mathematical objects. The internal language of toposes
  encapsulates and makes useful the following observation: The usual laws of
  reasoning apply to our favorite mathematical objects \emph{also in
  nontrivial, different ways}. I find this quite astonishing.

  The next slide shows a basic application of this train of thought in
  analysis. The presented example is of a rather simple nature and serves only
  to explain the translation process in an explicit and rigorous fashion,
  contrasting the previous slides which were somewhat short on details.
  Details on the application to homotopy theory can be found in the
  \fixedhref{https://homotopytypetheory.org/book/}{HoTT book}.}
\end{frame}

@@

    \item The approximative version of the intermediate value theorem (stating
    that for any~$\varepsilon > 0$, there is a number~$x$ such that~$|f(x)| <
    \varepsilon$)
    \fixedhref{https://mathoverflow.net/questions/253059/approximate-intermediate-value-theorem-in-pure-constructive-mathematics}{has
    a constructive proof} and therefore holds in~$\Sh(X)$. The
    external interpretation is that in continuous families of continuous
    functions, approximate zeros can locally be picked continuously.

@@

    \item Write~``$M^\Delta$'' for the set of all maps~$\Delta \to M$. A
    \emph{vector field} on a manifold (set)~$M$ is just a map~$X : M \to
    M^\Delta$ such that~$X(p)(0) = p$ for all~$p \in M$. A vector field~$X$
    induces an infinitesimal path~$\gamma : \Delta \to M^M$ in the space (set)
    of transformations of~$M$, by setting~$\gamma(\varepsilon) = (p \mapsto
    X(p)(\varepsilon))$.

@@

  Synthetic differential geometry is very well-developed and at your disposal.
  References include:
  \begin{itemize}
    \item \emph{\fixedhref{https://rawgit.com/iblech/mathezirkel-kurs/master/thema05-sdg/blatt05.pdf}{Synthetische
    Differentialgeometrie}}, notes for high school students (in German)
    \item \emph{\fixedhref{http://math.andrej.com/2008/08/13/intuitionistic-mathematics-for-physics/}{Intuitionistic
    mathematics for physics}}, a blog post by Andrej Bauer
    \item \emph{A Primer of Infinitesimal Analysis}, a short introduction by John Bell
    \item \emph{\fixedhref{http://home.imf.au.dk/kock/sdg99.pdf}{Synthetic
    Differential Geometry}}, the definitive book by Anders Kock
  \end{itemize}

@@

\section[Commutative algebra]{Applications in commutative algebra}

@@

    \item The ring~$A$ is a Prüfer domain if and only if~$A^\sim$ is a Bézout
    domain. Therefore any constructive theorem about Bézout domains entails a
    corresponding theorem about Prüfer domains. Bézout domains are quite rare,
    while Prüfer domains abound (for instance the ring of integers of any
    number field is a Prüfer domain, even constructively so).
  \end{itemize}

  There are
  \fixedhref{https://rawgit.com/iblech/internal-methods/master/slides-munich2018.pdf}{general
  metatheorems} which state a precise connection between~$A$ and~$A^\sim$,
  allowing in some cases to quickly pass from one to the other.}

@@

  Sometimes, the effects of this topos-theoretic reduction technique can be
  mimicked by classical techniques in commutative algebra such as passing
  to quotient rings or stalks. To the best of my knowledge, that's not the case
  for Grothendieck's generic freeness lemma. (In other cases, where they can,
  the classical techniques require the axiom of choice, while the
  topos-theoretic technique doesn't and therefore yields a more calculational,
  informative proof.)

  Details can be found
  \fixedhref{https://rawgit.com/iblech/internal-methods/master/slides-munich2018.pdf}{in
  this set of slides}.}

  \jnote{3}{As mentioned before, any proof involving the internal language can be
  \emph{unwound} to yield a direct proof not referencing toposes. Depending on
  the logical complexity of the statements appearing in the internal proof,
  this process can substantially lengthen the proof. In the case of the new
  proof of Grothendieck's generic freeness lemma, we were lucky to
  \fixedhref{https://arxiv.org/abs/1807.01231}{obtain a one-page proof} using
  this process. (That the resulting external proof was so short is because it
  was possible to eliminate the use of the Noetherian property from the
  internal proof; without that elimination, the unwound proof was much longer.)}
\end{frame}

@@

  \jnote{1}{One doesn't need to be an expert in topos theory in order to know
  that many notions in algebraic geometry are inspired by notions in algebra
  and that proofs in algebraic geometry often proceed by reducing to algebra.
  The internal language is a way of making this connection precise: In many
  cases, the former are simply interpretations of the latter internal
  to~$\Sh(X)$. Because this connection is precise instead of informal,
  additional value is gained: We can skip many basic proofs in algebraic
  geometry because they're just externalizations of proofs in algebra carried
  out internally to~$\Sh(X)$.}

  \jnote{2}{A basic example is as follows. A short exact sequence of sheaves of
  modules looks like a short exact sequence of plain modules from the internal
  point of view of~$\Sh(X)$. If the two outer sheaves are of finite type, then
  from the internal point of view, the two outer modules will look like
  finitely generated modules. Because the standard proof of the proposition
  quoted on the lower right is constructively valid, it follows that, from
  the internal point of view, the middle module is too finitely generated.
  Consulting the dictionary a second time, this amounts to saying that the
  middle sheaf is of finite type.

  A more advanced example is: The theorem ``any finitely generated vector space
  does \emph{not not} have a basis'' of constructive linear algebra entails, by
  interpretation in~$\Sh(X)$, that any sheaf of finite type over a reduced
  scheme is finite locally free on a \emph{dense open subset}.

  More details on this research program can be found in
  \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these
  notes}, partly reported on at the
  \fixedhref{https://www.youtube.com/watch?v=7S8--bIKaWQ}{2015 IHÉS topos
  theory conference}. Even though many important dictionary entries are still
  missing (for instance pertaining to derived categories and intersection
  theory), I believe that it is already in its current form useful to working
  algebraic geometers.

  The next slide illustrates a further, different way of approaching algebraic
  geometry using topos theory.}

@@

  \visible<2>{\hil{Synthetic approach:} model schemes \hil{directly as sets} in
  the internal universe of the \mbox{\hil{big Zariski topos}} of a base scheme.
  \small
  \[ \{ (x,y,z) \? (\affl)^3 \,|\, x^n+y^n-z^n=0 \} \]}

  \jnote{1}{At the
  \fixedhref{https://sbseminar.wordpress.com/2009/08/06/algebraic-geometry-without-prime-ideals/}{Secret
  Blogging Seminar}, there was an insightful long-running discussion on the
  merits of the two approaches. Two disadvantages of the approach using
  locally ringed spaces is that the underlying topological spaces don't
  actually parametrize ``honest'', ``geometric'' points, but the more complex
  notion of irreducible closed subsets; and that they don't work well in a
  constructive setting. (For this, they would have to be replaced by locally
  ringed locales.)

  The functorial approach is more economical, philosophically rewarding, and
  works constructively. Given a functor~$F : \Ring \to \Set$, we imagine~$F(A)$
  to be the set of~``$A$-valued points'' of the hypothetical scheme described
  by~$F$, the set of ``points with coordinates in~$A$''. These sets have direct
  geometric meaning. However, typically only field-valued points are easy
  to describe. For instance, the functor representing projective~$n$-space is
  given on fields by
  \[ \begin{array}{r@{}c@{}l}
  K &{}\longmapsto{}& \text{the set of lines through the origin in~$K^{n+1}$} \\
  && \qquad \cong \{ [x_0 \hg \cdots \hg x_n] \,|\, \text{$x_i \neq 0$ for some~$i$} \},
  \end{array} \]
  whereas on general rings it is given by
  \[ \begin{array}{r@{}c@{}l}
    A &{}\longmapsto{}& \text{the set of quotients~$A^{n+1} \twoheadrightarrow P$,
    where~$P$ is projective,} \\
    && \qquad \text{modulo isomorphism}.
  \end{array} \]
  It is these more general kinds of points which impart a sense of
  cohesion on the field-valued points, so they can't simply be dropped from
  consideration.\bigskip}

  \jnote{2}{This tension is resolved by observing that the category of
  functors~$\Ring \to \Set$ is a topos (the \emph{big Zariski topos} of~$\Spec(\ZZ)$)
  and that we can therefore employ its internal language. This language takes
  care of juggling stages behind the scenes. For instance, projective~$n$-space
  can be described by the naive expression
  \[ \{ (x_0,\ldots,x_n) : (\affl)^{n+1} \,|\, x_0 \neq 0 \vee \cdots \vee x_n
  \neq 0 \}/(\affl)^\times. \]
  This example illustrates the goal: to develop a synthetic account of
  algebraic geometry, in which schemes are plain sets and morphisms between
  schemes are maps between those sets. It turns out that there are many
  similarities with the well-developed synthetic account of differential
  geometry, but also important differences, and it also turns out that
  synthetic algebraic geometry has close connections to a certain age-old
  unresolved problem in topos theory, the \emph{mystery of nongeometric
  sequents}.

  Details are in
  \fixedhref{https://rawgit.com/iblech/internal-methods/master/slides-como2018.pdf}{this
  set of slides}.}
\end{frame}
